### 2011-02-06 Documentation/atropos/atropos.tex
ENH: fixed 'bug' in equation (4)
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@738 fd2da0a4-d968-48df-9cb5-d8c496287cbf
ENH:  Added ICM section (in red) and updated Atropos command line listing.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@762 fd2da0a4-d968-48df-9cb5-d8c496287cbf

 Atropos optimizes a class
of user-selectable objective functions each of which may
be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
requires {\em likelihood models} and {\em prior models} which enter
into Bayes' formula,
\begin{equation}\label{eq:bayes}
p(\mathbf{x}|\mathbf{y})=\underbrace{
p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{\mathbf{y}}, 
\end{equation} 
where the normalization term, $1/\mathbf{y}$, is a constant that does
not affect the optimization \citep{Sanjay-Gopal1998}.
Given choices for likelihood models and prior probabilities, the Bayesian
segmentation solution is the labeling $\hat{\mathbf{x}}$ which
maximizes the posterior probability, i.e.
\begin{align} \hat{\mathbf{x}} = \argmax_{\mathbf{x}}
\left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align} Similar to its predecessors, Atropos employs
the EM framework \citep{Dempster1977} to find maximum likelihood
solutions to this problem.  The following sections detail the 
Atropos EM along with choices for the likelihood and prior terms.

@@

\item[ICM Label Update (E-Step):] Given the initialization and fixed
model parameters, Atropos will evaluate the posterior probability for
each possible label and assign, to each voxel, the label that
maximizes the posterior.  
% {\bf \textcolor{red}{ FIXME more specific?}}.
\item[Parameter Update (M-Step):] 
Note that the posteriors used in the previous iteration are used to 
estimate the parameters at the current iteration.  
We use a common and elementary estimate of the mixing parameters:
\begin{align}
\gamma_k \leftarrow \frac{1}{N} \sum_{i=1}^N  p_k(l_k|y_i).
\end{align}
We update the parametric model parameters by computing, for each of
  $K$ labels, the mean,
\begin{align}  
 \mu_k \leftarrow \frac{ \sum_{i=1}^N  y_i p_k(l_k|y_i)}{ \sum_{i=1}^N p_k(l_k|y_i) },
\end{align}
and variance,
\begin{align}  
 \sigma^2_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k )^T
   p_k(l_k|y_i) (y_i - \mu_k )}{ \sum_{i=1}^N p_k(l_k|y_i) }.
\end{align}
The latter two quantities are modified, respectively, in the case of multivariate data as follows:
\begin{align}
 \boldsymbol{\mu}_k \leftarrow \frac{ \sum_{i=1}^N  \mathbf{y}_i p_k(l_k|\mathbf{y}_i)}{ \sum_{i=1}^N p_k(l_k|\mathbf{y}_i) },
\end{align}
and the $k^{th}$ covariance matrix, $\boldsymbol{\Sigma}_k$, is calculated from
\begin{align}
 \boldsymbol{\Sigma}_{k} \leftarrow \frac{ \sum_{i=1}^N  p_k(l_k|\mathbf{y}_i) ( \mathbf{y}_{i} - \boldsymbol{\mu}_{k} )^{\mathrm{T}}  ( \mathbf{y}_{i} - \boldsymbol{\mu}_{k} )}{1 - \sum_{i=1}^N p^2_k(l_k|\mathbf{y}_{i}) }.
\end{align}

This type of update is known as {\em soft} EM.  Hard EM, in
contrast, only uses sites containing label $l_k$ to update model $k$ parameters.
A similar pattern is used in non-parametric cases.
\end{description}

@@

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -a, --intensity-image [intensityImage,<adaptiveSmoothingWeight>]
     -b, --bspline [<numberOfLevels=6>,<initialMeshResolution=1x1x...>,<splineOrder=3>]
     -i, --initialization Random[numberOfClasses]
                          KMeans[numberOfClasses]
                          Otsu[numberOfClasses]
                          PriorProbabilityImages[numberOfClasses,
                            fileSeriesFormat(index=1 to numberOfClasses) or vectorImage,
                            priorWeighting,<priorProbabilityThreshold>]
                          PriorLabelImage[numberOfClasses,labelImage,priorWeighting]
     -x, --mask-image maskImageFilename
     -c, --convergence [<numberOfIterations=5>,<convergenceThreshold=0.001>]
     -k, --likelihood-model Gaussian
                            HistogramParzenWindows[<sigma=1.0>,<numberOfBins=32>]
     -m, --mrf [<smoothingFactor=0.3>,<radius=1x1x...>]
     -o, --output [classifiedImage,<posteriorProbabilityImageFileNameFormat>]
     -u, --minimize-memory-usage (0)/1
     -w, --winsorize-outliers BoxPlot[<lowerPercentile=0.25>,<upperPercentile=0.75>,
                               <whiskerLength=1.5>]
                              GrubbsRosner[<significanceLevel=0.05>,<winsorizingLevel=0.10>]
     -e, --use-euclidean-distance (0)/1
     -l, --label-propagation whichLabel[sigma=0.0,<boundaryProbability=1.0>]
     -h 
     --help 
\end{lstlisting} 
\end{command}

@@

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -i, --input-image inputImageFilename
     -x, --mask-image maskImageFilename
     -w, --weight-image weightImageFilename
     -s, --shrink-factor 1/2/3/4/...
     -c, --convergence [<numberOfIterations=50>,<convergenceThreshold=0.001>]
     -b, --bspline-fitting [splineDistance,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
                           [initialMeshResolution,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
     -t, --histogram-sharpening [<FWHM=0.15>,<wienerNoise=0.01>,<numberOfHistogramBins=200>]
     -o, --output [correctedImage,<biasField>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\doublespacing
\subsubsection{Integrating N4 Bias Correction}
Assumptions about bias correction may be thought of as another prior
model.  As such, the typical segmentation processing pipeline begins with an intensity
normalization/bias correction step using a method such as the recently developed N4
algorithm \citep{Tustison2010}.  N4 builds upon the popular
non-parametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
several methods proposed in the literature have taken an integrative
view of the segmentation problem by incorporating an intrinsic bias
correction step into the actual workflow.  The
advancements introduced with N4 permit such an adaptive
integration with Atropos.  Recent demonstrations suggest improved
white matter segmentation produces better gain field estimates using
N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
may opt to use, for instance, the posterior probability map of white matter at the current
iteration as a weighted mask for input to N4.  This is done by setting
the `{\ttfamily --weight-image}' option on the N4 command line call
(see Listing \ref{listing:n4}) to the posterior probability image
corresponding to the white matter produced as output in the Atropos
call, i.e. `{\ttfamily Atropos --output}'.  N4 is
currently being added as a standard part of the Insight Toolkit.  The
evaluation section will illustrate inclusion of Atropos, N4 and ANTs
in a brain processing pipeline.

>>>>

 Atropos optimizes a class
of user-selectable objective functions each of which may
be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
requires {\em likelihood models} and {\em prior models} which enter
into Bayes' formula,
\begin{equation}\label{eq:bayes}
p(\mathbf{x}|\mathbf{y})=\underbrace{
p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{p(\mathbf{y})}, 
\end{equation} 
where the normalization term, $1/\mathbf{y}$, is a constant that does
not affect the optimization \citep{Sanjay-Gopal1998}.
Given choices for likelihood models and prior probabilities, the Bayesian
segmentation solution is the labeling $\hat{\mathbf{x}}$ which
maximizes the posterior probability, i.e.
\begin{align} \hat{\mathbf{x}} = \argmax_{\mathbf{x}}
\left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align} Similar to its predecessors, Atropos employs
the EM framework \citep{Dempster1977} to find maximum likelihood
solutions to this problem.  The following sections detail the 
Atropos EM along with choices for the likelihood and prior terms.

@@

\item[Label Update (E-Step):] Given the initialization and fixed
model parameters, {\color{red}{Atropos is capable of updating the current label 
estimates using either a {\em synchronous} or {\em asynchronous} scheme.
The former is characterized by iterating through the image and determining
which label maximizes the posterior probability without updating any
labels until all voxels in the mask have been visited at which point
all the voxel labels are updated simultaneously (hence the descriptor 
``synchronous'').  This option is specified with \texttt{--icm [0]}. 
However, unlike asynchronous schemes characteristic of ICM, synchronous 
updates lack convergence guarantees.  To determine the labeling 
which maximizes the posterior probability for the asynchronous option, 
an ``ICM code'' image is created once for all iterations by iterating 
through the image and assigning an ICM code label to each voxel in the 
mask such that each MRF neighborhood has a non-repeating code label set.  
Thus each masked voxel in the ICM code image is assigned a value in the 
range $\{1,\ldots,C\}$ where $C$ is the maximum code label.
Such an image can be created and viewed with Atropos by assigning a valid
filename in the \texttt{--icm [1]} set of options.  An example is given in 
Figure \ref{fig:r16icm}.  For each iteration, a random ordering of the
set $\{1,\ldots,C\}$ is created which prescribes the order in which the
voxel labels are updated asynchronously. After the first pass through the
set of code labels, additional passes can further increase the posterior
probability until convergence (in $\sim$5 iterations).  One can specify
a maximum number of these ``ICM iterations'' on the command line.}}

\begin{figure}
\begin{center}
\begin{tabular}{cc}
\includegraphics[width=70mm]{r16slice.png} &
\includegraphics[width=70mm]{r16icm.png} \\
(a) & (b) \\
\end{tabular}
\end{center}
\caption{\baselineskip 12pt \small  (a) An adult brain image slice and (b) the corresponding ICM code image created using a $5\times5$ MRF neighborhood with a maximum ICM code label of `13'.  }
\label{fig:r16icm}
\end{figure}

% {\bf \textcolor{red}{ FIXME more specific?}}.
\item[Parameter Update (M-Step):] 
Note that the posteriors used in the previous iteration are used to 
estimate the parameters at the current iteration.  
We use a common and elementary estimate of the mixing parameters:
\begin{align}
\gamma_k \leftarrow \frac{1}{N} \sum_{i=1}^N  p_k(l_k|y_i).
\end{align}
We update the parametric model parameters by computing, for each of
  $K$ labels, the mean,
\begin{align}  
 \mu_k \leftarrow \frac{ \sum_{i=1}^N  y_i p_k(l_k|y_i)}{ \sum_{i=1}^N p_k(l_k|y_i) },
\end{align}
and variance,
\begin{align}  
 \sigma^2_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k )^T
   p_k(l_k|y_i) (y_i - \mu_k )}{ \sum_{i=1}^N p_k(l_k|y_i) }.
\end{align}
The latter two quantities are modified, respectively, in the case of multivariate data as follows:
\begin{align}
 \boldsymbol{\mu}_k \leftarrow \frac{ \sum_{i=1}^N  \mathbf{y}_i p_k(l_k|\mathbf{y}_i)}{ \sum_{i=1}^N p_k(l_k|\mathbf{y}_i) },
\end{align}
and the $k^{th}$ covariance matrix, $\boldsymbol{\Sigma}_k$, is calculated from
\begin{align}
 \boldsymbol{\Sigma}_{k} \leftarrow \frac{ \sum_{i=1}^N  p_k(l_k|\mathbf{y}_i) ( \mathbf{y}_{i} - \boldsymbol{\mu}_{k} )^{\mathrm{T}}  ( \mathbf{y}_{i} - \boldsymbol{\mu}_{k} )}{1 - \sum_{i=1}^N p^2_k(l_k|\mathbf{y}_{i}) }.
\end{align}
This type of update is known as {\em soft} EM.  Hard EM, in
contrast, only uses sites containing label $l_k$ to update model $k$ parameters.
A similar pattern is used in non-parametric cases.
\end{description}

@@

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -a, --intensity-image [intensityImage,<adaptiveSmoothingWeight>]
     -b, --bspline [<numberOfLevels=6>,<initialMeshResolution=1x1x...>,<splineOrder=3>]
     -i, --initialization Random[numberOfClasses]
                          KMeans[numberOfClasses]
                          Otsu[numberOfClasses]
                          PriorProbabilityImages[numberOfClasses,
                            fileSeriesFormat(index=1 to numberOfClasses) or vectorImage,
                            priorWeighting,<priorProbabilityThreshold>]
                          PriorLabelImage[numberOfClasses,labelImage,priorWeighting]
     -p, --posterior-formulation Socrates[<useMixtureModelProportions=1>,
                                   <initialAnnealingTemperature=1>,<annealingRate=1>,
                                   <minimumTemperature=0.1>]
                                 Plato[<useMixtureModelProportions=1>,
                                   <initialAnnealingTemperature=1>,<annealingRate=1>,
                                   <minimumTemperature=0.1>]
     -x, --mask-image maskImageFilename
     -c, --convergence [<numberOfIterations=5>,<convergenceThreshold=0.001>]
     -k, --likelihood-model Gaussian
                            HistogramParzenWindows[<sigma=1.0>,<numberOfBins=32>]
     -m, --mrf [<smoothingFactor=0.3>,<radius=1x1x...>]
     -g, --icm [<useAsynchronousUpdate=1>,<maximumNumberOfICMIterations=1>,
                <icmCodeImage=`'>]     
     -o, --output [classifiedImage,<posteriorProbabilityImageFileNameFormat>]
     -u, --minimize-memory-usage (0)/1
     -w, --winsorize-outliers BoxPlot[<lowerPercentile=0.25>,<upperPercentile=0.75>,
                                <whiskerLength=1.5>]
                              GrubbsRosner[<significanceLevel=0.05>,<winsorizingLevel=0.10>]
     -e, --use-euclidean-distance (0)/1
     -l, --label-propagation whichLabel[sigma=0.0,<boundaryProbability=1.0>]
     -h 
     --help 
\end{lstlisting} 
\end{command}

@@

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -i, --input-image inputImageFilename
     -x, --mask-image maskImageFilename
     -w, --weight-image weightImageFilename
     -s, --shrink-factor 1/2/3/4/...
     -c, --convergence [<numberOfIterations=50>,<convergenceThreshold=0.001>]
     -b, --bspline-fitting [splineDistance,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
                           [initialMeshResolution,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
     -t, --histogram-sharpening [<FWHM=0.15>,<wienerNoise=0.01>,<numberOfHistogramBins=200>]
     -o, --output [correctedImage,<biasField>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\doublespacing
\subsubsection{Integrating N4 Bias Correction}
Assumptions about bias correction may be thought of as another prior
model.  As such, the typical segmentation processing pipeline begins with an intensity
normalization/bias correction step using a method such as the recently developed N4
algorithm \citep{Tustison2010}.  N4 builds upon the popular
non-parametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
several methods proposed in the literature have taken an integrative
view of the segmentation problem by incorporating an intrinsic bias
correction step into the actual workflow.  The
advancements introduced with N4 permit such an adaptive
integration with Atropos.  Recent demonstrations suggest improved
white matter segmentation produces better gain field estimates using
N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
may opt to use, for instance, the posterior probability map of white matter at the current
iteration as a weighted mask for input to N4.  This is done by setting
the `{\ttfamily --weight-image}' option on the N4 command line call
(see Listing \ref{listing:n4}) to the posterior probability image
corresponding to the white matter produced as output in the Atropos
call, i.e. `{\ttfamily Atropos --output}'.  N4 was recently added  
to the Insight Toolkit repository%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1N4MRIBiasFieldCorrectionImageFilter.html
} 
where it is built and tested on
multiple platforms nightly.  The
evaluation section will illustrate inclusion of Atropos, N4 and ANTs
in a brain processing pipeline.
